---
title: "Poisson Random Variables"
author: "Advanced Statistics"
output: beamer_presentation
---

## Two Goals a Game... on average

A soccer team scores two goals per game, what is the probability that they will score:
\begin{itemize}
\item 0 goals
\item 1 goal
\item 2 goals
\item 3 goals
\item ... 10 goals?
\end{itemize}


## Can we solve this using the binomial formula?

Let's **pretend** (we'll return to whether this is reasonable in a minute) that this team has 5 trials or attempts a game and has a 40% chance of success in each trial.

$n = 5, p = 0.4$
$X =$ # of goals scored (a random variable)
$E[X] = np = 2$ goals

$P(X = k) = {n \choose k} p^k (1-p)^{n-k}$

So,

$P(X = 0) = {5 \choose 0} .4^0 (1-.4)^{5-0} = 0.6^5 =$ ```r .6^5```
$P(X = 1) = {5 \choose 1} .4^1 (1-.4)^{5-1} =$ ```r dbinom(1, 5, .4)```

## The Full Distribution of Probabilities given the 5 attempt assumption

```{r, echo=FALSE}
plot(0:5, dbinom(0:5, 5, .4), type="h", col="blue", xlab="Number of Goals", ylab="Probability")
text(0:5, dbinom(0:5, 5, .4), dbinom(0:5, 5, .4))

```

## But what if they have 10 attempts?

With 10 attempts 

$n = 10$
$p = 0.2$
$E[X] = np = 2$ goals (just as before)

$P(X = 1) = {10 \choose 1} .2^1 (1-.2)^{1-1} =$ ```r dbinom(1, 10, .2)```

```{r, echo=FALSE}
plot(0:5-0.1, dbinom(0:5, 5, .4), type="h", col="blue", xlab="Number of Goals", ylab="Probability", xlim=c(-0.5,6))
text(0:5-0.1, dbinom(0:5, 5, .4), round(dbinom(0:5, 5, .4),3))

points(0:10+.1, dbinom(0:10, 10, .2), type="h", col="red", xlab="Number of Goals", ylab="Probability")
text(0:10+.1, dbinom(0:10, 10, .2), round(dbinom(0:10, 10, .2),3))
```

## But what if they have a large number of attempts?

Blue: $n = 50, p = .04, E[X] = np = 2$

Red: $n = 500, p = .004, E[X] = np = 2$

```{r, echo=FALSE}
n <- 50
pg <- 2/n
maxk <- 8
plot(0:maxk-0.1, dbinom(0:maxk, n, pg), type="h", col="blue", xlab="Number of Goals", ylab="Probability", xlim=c(-0.5,8.5))
text(0:maxk-0.25, dbinom(0:maxk, n, pg), round(dbinom(0:maxk, n, pg),3))

n <- 500
pg <- 2/n
points(0:maxk+.1, dbinom(0:maxk, n, pg), type="h", col="red", xlab="Number of Goals", ylab="Probability")
text(0:maxk+.25, dbinom(0:maxk, n, pg), round(dbinom(0:maxk, n, pg),3))
```

## Letting the Number of Attempts go to Infinity

Instead of some number of attempts, $n,$ and rate of success, $p,$ we'll think about an expected number of successes $\lambda$ and let n, the number of attempts, go to infinity (and p go to zero).

$P(X = k) = {n \choose k} p^k (1-p)^{n-k}$

$p = \frac{\lambda}{n}$

$P(X = k) = \lim_{n \to \infty} {n \choose k} (\frac{\lambda}{n})^k (1-\frac{\lambda}{n})^{n-k}$


## What does this distribution look like?


$P(X = k) = \lim_{n \to \infty} {n \choose k} (\frac{\lambda}{n})^k (1-\frac{\lambda}{n})^{n-k}$


Can be simplified  to:

$P(X = k)= \frac{\lambda^k }{k!} e^{-\lambda}$

## Back to our soccer calculation:
$P(X = k)= \frac{\lambda^k }{k!} e^{-\lambda}$

$\lambda = 2$

$P(X = 0)= \frac{2^0 }{0!} e^{-2} = e^{-2}$ = ```r round(exp(-2),3)```

$P(X = 1)= \frac{2^1 }{1!} e^{-2} = 2 e^{-2}$ = ```r round(2*exp(-2),3)```

$P(X = 2)= \frac{2^2 }{2!} e^{-2} =\frac{4}{2} e^{-2}$ = ```r round(2*exp(-2),3)```

$P(X = 3)= \frac{2^3 }{3!} e^{-2} = \frac{8}{6} e^{-2}$ = ```r round((4/3)*exp(-2),3)```

$P(X = 4)= \frac{2^4 }{4!} e^{-2} = \frac{16}{24} e^{-2}$ = ```r round((2/3)*exp(-2),3)```

$P(X = 5)= \frac{2^5 }{5!} e^{-2} = \frac{32}{120} e^{-2}$ = ```r round((32/120)*exp(-2),3)```

## When does this formula work?
$P(X = k)= \frac{\lambda^k }{k!} e^{-\lambda}$

For Rare Events when $p$ is very small or $n >> k$

*and*

rate of successes is constant

Soccer goals and Hockey goals are “rare” relative to the number of plays in the game.

I could also use this distribution to model the number of times that I win the Hamilton lottery in the next year or decade, the number of earthquakes (of a given magnitude in next century), the number of typos in a book or the number of mutations on a DNA strand.  







