---
title: "Poisson Formula Derivation"
author: "Advanced Statistics"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

## Letting the Number of Attempts go to Infinity

Instead of some number of attempts, $n,$ and rate of success, $p,$ we'll think about an expected number of successes $\lambda$ and let n, the number of attempts, go to infinity (and p go to zero).

$P(X = k) = {n \choose k} p^k (1-p)^{n-k}$

$p = \frac{\lambda}{n}$

$P(X = k) = \lim_{n \to \infty} {n \choose k} (\frac{\lambda}{n})^k (1-\frac{\lambda}{n})^{n-k}$


## What does this distribution look like?


$P(X = k) = \lim_{n \to \infty} {n \choose k} (\frac{\lambda}{n})^k (1-\frac{\lambda}{n})^{n-k}$


We'll rewrite this as:

$P(X = k) = \frac{\lambda^k}{k!} \cdot   [k! \cdot {n \choose k} (\frac{1}{n})^k] \cdot [(1-\frac{\lambda}{n})^{n}]\cdot [(1-\frac{\lambda}{n})^{-k}]$

and then break this into four parts:

$P(X = k) = \frac{\lambda^k}{k!} \cdot   A \cdot B \cdot C$

where $A =\lim_{n \to \infty} k! \cdot {n \choose k} (\frac{1}{n})^k$ and $B = \lim_{n \to \infty} (1-\frac{\lambda}{n})^{n}$ and $C = \lim_{n \to \infty} (1-\frac{\lambda}{n})^{-k}$





## Starting with A:

$A = \lim_{n \to \infty} k! \cdot {n \choose k} (\frac{1}{n})^k$

$\lim_{n \to \infty} k! \cdot \frac{n!}{k! (n-k)!} (\frac{1}{n})^k$

$\lim_{n \to \infty} \frac{n!}{(n-k)!} (\frac{1}{n})^k$

$\lim_{n \to \infty} n \cdot (n-1) \cdot (n-2) \cdot ... \cdot (n-k+1) (\frac{1}{n})^k$

$\lim_{n \to \infty} \frac{n \cdot (n-1) \cdot (n-2) \cdot ... \cdot (n-k+1)}{n^k}$

$\lim_{n \to \infty} \frac{n}{n} \cdot \frac{n-1}{n} \cdot \frac{n-2}{n} \cdot ... \cdot \frac{n-k+1}{n} = 1 \cdot 1 \cdot 1 \cdot ... \cdot 1 = 1$

So A just equals 1.

## Now part B:


$B = \lim_{n \to \infty} (1-\frac{\lambda}{n})^n$

subtitute $j = -1 \cdot \frac{n}{\lambda}$ and we have

$B = \lim_{j \to \infty} (1+\frac{1}{j})^{-1 \cdot \lambda j} = \lim_{j \to \infty} [(1+\frac{1}{j})^{j}]^{-1 \cdot \lambda}$

But what is $\lim_{j \to \infty} (1+\frac{1}{j})^{j}$ ?

## Euler's Number

```{r, echo=FALSE}
f <- function(x){
  (1+1/x)^x
}

vals = seq(0, 8, 0.1)

x <- 10^vals

library(ggplot2)
qplot(x, f(x))+
  scale_x_log10()+
  ylab("(1+1/x)^x")+
  geom_hline(yintercept=exp(1), color="red")
```


## Finally, part C:

So $B = e^{-\lambda}$

What about C?  Recall, $C = \lim_{n \to \infty} (1-\frac{\lambda}{n})^{-k}$.

and as $n \to \infty$ this becomes $1^{-k}$ or $\frac{1}{1^k}$ or just 1.

Putting it all together we have:

$P(X = k) = \frac{\lambda^k}{k!} \cdot   1 \cdot e^{-\lambda} \cdot 1$

$P(X = k)= \frac{\lambda^k }{k!} e^{-\lambda}$
